---
title: "Brewed Behavior: An analysis of Coffee Consumption"
subtitle: "Proposal"
format: html
editor: visual
---

```{r load-library, warning=FALSE}
#| label: load-pkgs
#| message: false
library(tidyverse)
```

## Dataset

```{r load-dataset}
#| label: load-dataset
#| message: false

coffee <- read_csv("data/coffee_survey.csv")
```

## **Introduction**

When selecting our dataset, our team was interested in two key considerations: relatability and real-world relevance. We sought a dataset centered on a topic that most people could readily understand and connect with, while also providing an opportunity to explore a practical, real-world question. This dataset particularly appealed to us because it focuses on something familiar to millions of people, particularly in the United States: coffee. We were especially interested in examining the financial aspects of coffee consumption, with coffee shop owners and roasters as our intended audience. Specifically, we wanted to understand who spends the most on coffee each month and who is willing to pay a premium for a single cup. Such information may be relevant to business decisions involving pricing, subscription models, premium offerings, and customer targeting.

The dataset was collected through a one-time, voluntary survey; therefore, our analysis is primarily descriptive and exploratory, with a focus on identifying patterns and associations. Establishing causality is beyond the scope of what this dataset can support. An earlier analysis of this survey by McKeon Aloe (2023) compared coffee spending across roast preferences and income levels and observed that people who spent more tended to prefer lighter roasts and that the most people would pay for a cup varied little by income. However, less is known about the characteristics associated with higher monthly coffee expenditures and a greater willingness to pay per cup. Our goal is to explore whether these outcomes are associated with factors such as coffee consumption, level of interest in coffee, preference for stronger coffee, and demographic characteristics.

## **Dataset description**

Our dataset is based on the Great American Coffee Taste Test, conducted in October 2023 by James Hoffmann, an English barista, YouTuber, and author, in collaboration with the coffee company Cometeer. The data are available through the TidyTuesday project (May 14, 2024). Participants in the United States purchased a kit containing four coffees, tasted them blind during a livestream, and completed an online survey about their coffee habits, spending, and demographic characteristics, along with their ratings of the four coffees.

The resulting dataset contains `r format(nrow(coffee), big.mark = ",")` rows, with each row representing a survey submission, and `r ncol(coffee)` variables. Most variables are categorical and stored as text, including binned ranges for age and spending. The numeric variables include self-rated coffee expertise, measured on a 1-10 scale, and ratings of bitterness, acidity, and personal preference for each of the four coffees, measured on a 1-5 scale. The sample is likely to consist largely of people who are engaged with coffee because participation was voluntary and the survey was associated with a specialty-coffee YouTuber. Therefore, the findings are not generalizable and should not be interpreted as representative of the general U.S. population.

## **Dataset Details**

### **Respondents**

We first explored the dataset to provide context about the population represented in the survey. In the dataset, 50% of respondents are between 25 and 34 years old, and 72% reported their gender are men. Their coffee-related behaviors also suggest a strong interest in coffee. Approximately 92% drink coffee at home; among the 3,657 home brewers, 63% use pour-over and 42% use espresso. In addition, 66% drink their coffee black, while 43% add milk, a dairy alternative, or creamer, and 13% add sugar or another sweetener. Among the 710 respondents who reported purchasing coffee on the go, 62% purchase from specialty coffee shops, 55% from local cafés, and 46% from national chains. Most respondents reported drinking coffee because they enjoy its taste (94%), while 57% reported drinking it for the caffeine. The level of engagement with coffee is also reflected in spending on equipment: 22% of respondents reported spending more than \$1,000 on coffee equipment during the previous five years.

### **Variables used for analysis**

To answer our research questions, we selected variables that capture coffee consumption, coffee preferences and expertise, spending, willingness to pay, and demographic characteristics.

| Variable | Description | Type | Question | Missing (%) |
|---------------|---------------|---------------|---------------|---------------|
| cups | Cups of coffee consumed per day | Ordered categories | Q1 | 2.3 |
| strength | Preferred coffee strength | Ordered categories | Q1 | 3.1 |
| expertise | Self-rated coffee expertise | Numeric (1-10) | Q1 | 2.6 |
| total_spend | Typical monthly spending on coffee | Ordered categories (dollars) | Q1 | 13.1 |
| most_willing | Maximum amount willing to pay for a single cup | Ordered categories (dollars) | Q1, Q2 | 13.2 |
| age | Age group | Ordered categories | Q2 | 0.8 |
| gender | Gender | Categorical | Q2 | 12.8 |
| employment_status | Employment status | Categorical | Q2 | 15.4 |

### **Data considerations and limitations**

As aforementioned, the sample is self-selected and consists primarily of engaged coffee consumers, so findings should not be generalized to the broader U.S. coffee-drinking population. Approximately 13% of respondents are missing both spending variables, which may reflect survey drop-off and could result in slight overrepresentation of more engaged coffee consumers among those included in the analysis. Spending and willingness to pay are recorded as ranges stored as text, thus, we will convert them to ordered factors and treat them as categories. Some demographic groups are also small and may need to be combined for analysis. Many heavy drinkers in this sample also rate themselves as coffee experts, so the groups we compare in Question 1 overlap. Our plots will show each characteristic separately, so a pattern in one group may partly reflect another.

## **Questions**

### **Question 1**

We want to investigate whether coffee consumption, preferred coffee strength, and coffee expertise are associated with how much respondents spend on coffee each month and how much they are willing to pay for a single cup. This will help us understand which coffee-related characteristics are associated with higher spending and willingness to pay.

### **Question 2**

We want to investigate whether willingness to pay for a single cup of coffee varies across age, gender, and employment status. This will allow us to examine whether demographic characteristics are associated with differences in the maximum amount respondents are willing to pay.

## **Analysis Plan**

### **Question 1: Who spends more on coffee each month, and who will pay more for a single cup: heavy drinkers, strong-coffee drinkers, or coffee enthusiasts?**

Coffee consumption can vary substantially between individuals, from the amount of coffee they drink to how strongly they prefer it and how knowledgeable they consider themselves about coffee. These characteristics may be related to how much people spend on coffee each month and how much they are willing to pay for an individual cup. Therefore, we will examine whether heavier coffee consumption, stronger coffee preferences, and greater coffee expertise are associated with higher spending and willingness to pay.

We will analyze the relationships between cups, strength, and expertise and the two spending-related variables, total_spend and most_willing. Because total_spend and most_willing are reported as ordered dollar ranges we will have to clean the data. We are going to use the maximum of this range because people consistently underestimate the amount of money that they spend on coffee. Coffee purchases are underestimated due to mental accounting, where small, habitual daily expenses are mentally classified as negligible rather than part of a strict budget. Our first plot will compare monthly coffee spending across categories of coffee consumption such as caffeine amount, roast, and coffee expertise. Our second plot will examine willingness to pay for a single cup across these same characteristics

### **Question 2: Is willingness to pay for a cup of coffee associated with age, gender, and employment status?** 

Coffee spending and willingness to pay may also differ across demographic groups. Understanding how willingness to pay varies across age, gender, and employment status can provide additional context for the types of respondents represented in the survey and reveal whether willingness to pay differs among demographic groups.

We will examine the relationship between most_willing and each of the three demographic variables: age, gender, and employment status. We will use the maximum amount respondents reported being willing to pay for a single cup, represented by the most_willing range, to compare willingness to pay across demographic groups. Because the responses are recorded as dollar ranges, we will treat most_willing as an ordered categorical variable rather than assigning respondents an exact dollar amount. Our two plots will show how the distribution of maximum willingness to pay varies across the selected demographic groups.

## Discussion

We expect self-rated coffee expertise to be a stronger predictor of both monthly spending and willingness to pay per cup than raw consumption (cups per day) or preferred strength, since respondents who rate themselves as more knowledgeable about coffee are more likely to seek out premium or specialty options regardless of how much coffee they drink. For Question 2, we anticipate that willingness to pay will vary most clearly across employment status, given likely differences in disposable income, and moderately across age, particularly if younger respondents favor specialty or trend-driven purchases while older respondents show more consistent spending habits. We do not expect a strong gender effect but will explore it given its inclusion in the dataset. The primary anticipated challenge is the overlap between heavy drinkers, strong-coffee drinkers, and self-rated experts, which may make it difficult to isolate which factor most directly drives spending; we will address this by visualizing each characteristic separately rather than combining them into a single model. We also anticipate some loss of precision due to missing spending data (\~13%) and the binned nature of the spending and willingness-to-pay variables. A likely success of this analysis will be a clear, practical takeaway for coffee shop owners and roasters regarding which customer segments are most worth targeting with premium offerings or subscription models.

## Timeline

- **Sept 17 - Sept 21:** Explore dataset and codebook individually; each member drafts two candidate research questions; complete initial descriptive/exploratory analysis.

- **Sept 22 - Sept 25:** Finalize the two team research questions; draft assigned proposal sections (Introduction, Dataset, Analysis Plan, Discussion, Timeline) collaboratively; target a solid initial draft by Friday, Sept 25.

- **Sept 26 - Sept 28:** Review the full draft as a team, revise, incorporate feedback, transfer finalized sections from the Google Doc into proposal.qmd in RStudio, and commit/push to GitHub.

- **Sept 29 - Oct 5:** Begin data cleaning and exploratory analysis for Question 1 and Question 2; draft initial visualizations.

- **Oct 6 - Oct 10:** Finalize plots (two per question), write up Results and interpretation for each question.

- **Oct 11 – Oct 13:** Full team review of the complete project - analysis, writing, and plots; proofread and polish.

- **Oct 14:** Submit final Project 1 by 11:59 pm..
